Fractional Types

### 2013-02-24 frac.tex
minor things

Historically, versions of compact closed categories were introduced by
Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the generalization of
monoidal categories to model quantum computation.  The first and most
intuitive example of such categories is the category of finite sets and
relations. In that case, the dual operation (i.e., the fractional type)
collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
interpretation of fractionals, this is still an interesting category as it
provides a model for a reversible higher-order language. Indeed, any relation
is reversible via the standard \emph{converse} operation on a relation.
Furthermore, as any relation between {{b1}} and {{b2}} can be represented as
a subset of {{b1 * b2}}, i.e., as a value.

@@

The semantics above defines a relation rather than a function. It is still
reversible in the sense that it ``mode-checks''. In other words, for any
combinator {{c}} we can treat either its left or its right side as an input;
this allows us to define {{sym}} as just relational converse, which amounts
to an argument swap. We can then give an operational interpretation to the
semantic relation by defining two functions: a forward evaluator and a
backwards evaluator that exercise the relation in opposite directions, but
which now can return a \emph{set} of results. This operational view exploits
the usual conversion of a relation from a subset of {{A * B}} to a function
from {{A}} to the powerset of {{B}} whose composition is given by the Kleisli
composition in the powerset monad. It is straightforward to check that if the
forward evaluation of {{c v1}} produces {{v2}} as a possible answer then the
backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.  One
can check that all relations for {{Pi}} are total functional relations whose
converse are also total functional relations -- aka isomorphisms.  However
{{eta*_b}} relates {{()}} to multiple values (one for each value of {{b}}),
and is thus not functional; {{eps*_b}} is exactly its relational converse, as
expected.

@@

For small examples, finding such combinators by trial and error is a
relatively straightforward but tedious task. It is, however, possible to
automate this task by expressing what is essentially a reversible SAT solver
as follows. In the usual setting, an instance of SAT is a function~{{f}}
which, when given some boolean inputs, returns {{true}} or {{false}}. The
function returns {{true}} when the inputs satisfy the constraints imposed by
the structure of {{f}} and a solution to the SAT problem is the set of all
inputs on which {{f}} produces {{true}}. The basic idea of our construction
is to use {{trace*}} to annihilate values that fail to satisfy the
constraints represented by the SAT instance {{f}}.

>>>>

Historically, versions of compact closed categories were introduced by
Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the generalization of
monoidal categories to model quantum computation.  The first and most
intuitive example of such categories is the category of finite sets and
relations. In that case, the dual operation (i.e., the fractional type)
collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
interpretation of fractionals, this is still an interesting category as it
provides a model for a reversible higher-order language. Indeed, any relation
is reversible via the standard \emph{converse} operation on a relation.
Furthermore, any relation between {{b1}} and {{b2}} can be represented as
a subset of {{b1 * b2}}, i.e., as a value.

@@

The semantics above defines a relation rather than a function. It is still
reversible in the sense that it ``mode-checks''. In other words, for any
combinator~{{c}} we can treat either its left or its right side as an input;
this allows us to define {{sym}} as just relational converse, which amounts
to an argument swap. We can then give an operational interpretation to the
semantic relation by defining two interpreters: a forward evaluator and a
backwards evaluator that exercise the relation in opposite directions, but
which now can return a \emph{set} of results. This operational view exploits
the usual conversion of a relation from a subset of {{A * B}} to a function
from {{A}} to the powerset of {{B}} whose composition is given by the Kleisli
composition in the powerset monad. It is straightforward to check that if the
forward evaluation of {{c v1}} produces {{v2}} as a possible answer then the
backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.  One
can check that all relations for {{Pi}} are total functional relations whose
converse are also total functional relations -- aka isomorphisms.  However
{{eta*_b}} relates {{()}} to multiple values (one for each value of {{b}}),
and is thus not functional; {{eps*_b}} is exactly its relational converse, as
expected.

@@

For small examples, finding such combinators by trial and error is a
relatively straightforward but tedious task. It is, however, possible to
automate this task by expressing what is essentially a reversible SAT solver
as follows. In the usual setting, an instance of SAT is a function~{{f}}
which, when given some boolean inputs, returns {{true}} or {{false}}. The
function returns {{true}} when the inputs satisfy the constraints imposed by
the structure of {{f}} and a solution to the SAT problem is the set of all
inputs on which {{f}} produces {{true}}. The basic idea of our construction
is to use {{trace*}} to annihilate values that fail to satisfy the
constraints represented by the SAT instance {{f}}. (The accompanying code
details the construction of such a solver.)
